Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-21 paper/paper.tex
Revise section 3 (WIP)

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Themetavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }

The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.
\scw{Spell this out a bit more about what goes wrong. Where does the
proof break? Also spell out how this enforces predicativity.}

To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$.

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\scw{I didn't understand this figure without reading the text below.}
\end{figure}

With the definition from Figure~\ref{fig:logrelrec}, we can
show that the same introduction rules for $[[Interp I i A S]]$ are
admissible for $[[InterpR i A S]]$, by
instantiating $[[I]]$ with $[[I i]] := [[{A | exists S , InterpR i A
  S}]]$, the same function $[[I]]$ used in the definition of $[[InterpR i A
S]]$. We give as example the following rules.
\begin{center}
\drule[]{IR-Void} \qquad \drule[]{IR-Set}
\end{center}

@@

In the rest of the section, we establish some important facts about
the logical relation that will be useful for proving fundamental theorem
in Section~\ref{sec:logrelproof}.

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.\scw{Put footnotes on lemma name, or on (1), (2), etc}
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}
  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}}
  \end{enumerate}
\end{lemma}
\begin{proof}
  \scw{Why is this proof interesting and worth typesetting? Should
    it be in an appendix so that it doesn't interrupt the flow of
    reading?}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    These properties are exactly what's\scw{avoid contractions} required in the conclusion.
  \end{description}
\end{proof}

@@

property. The following property shows that preservation holds in the
usual forward direction too.

@@

>>>>

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }

The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of $<$ relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$.

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
The definition of in Figure~\ref{fig:logrelrec} explains how the $[[j
< i]]$ constraint in \rref{I-Set} makes our system predicative; the
interpretation of the $[[i]]_{th}$ universe is only dependent on
universes at strictly lower than $[[i]]$, which have been defined earlier.
Removing the ordering constraint would result in a
system where one can encode Girard's
paradox~\citep{girard-thesis}. Our proof would no longer work with the
restriction removed since the definition of $[[InterpR i A S]]$ would
not be well-founded when $[[Interp I i A S]]$ calls $[[I]]$ on
universe levels greater than or equal to $[[i]]$, which are yet to be defined.

By unfolding the definition in Figure~\ref{fig:logrelrec}, we
show that the same introduction rules for $[[Interp I i A S]]$ are
admissible for $[[InterpR i A S]]$, by
instantiating $[[I]]$ with $[[I i]] := [[{A | exists S , InterpR i A
  S}]]$, the same function $[[I]]$ used in the definition of $[[InterpR i A
S]]$. We give as example the following rules.
\begin{center}
\drule[]{IR-Void} \qquad \drule[]{IR-Set}
\end{center}

@@

In the rest of the section, we prove four important facts about
our logical relation: irrelevance (Lemma~\ref{lemma:logrelcoherence}),
functionality (Lemma~\ref{lemma:logreldeter}), cumulativity
(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure
property (Lemma~\ref{lemma:logrelNbackclos}).

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi A0
B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    These properties are exactly the ones from the conclusion.
  \end{description}
\end{proof}

@@

property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
$[[Interp I i B S]]$ also holds. The following property shows that
preservation holds in the usual forward direction too.

@@

\appendix

\subsection{More detailed proofs}
\begin{lemma}[Preservation of the logical relation]
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.
  \begin{description}
  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
    be $[[Void]]$ and the result trivially follows.
  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
      refl , a <=> b  }]]$ and, by inverting the derivation of
    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
    <=> b0]]$. The result then immediately follows from the fact that
    coherence is an equivalence relation.
  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
    such that:
    \begin{itemize}
    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
        a #) }]]$
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
      a #) }]]$. By \rref{I-Pi}, it suffices to show:
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \end{itemize}
    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
    conditions follow immediately from the induction hypothesis.
  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
    property of parallel
    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
    B0 S]]$. By \rref{I-Red} and
    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
  \end{description}
\end{proof}
